---
title: "Benchmarking hmmfb"
author: "Nicholas Knoblauch"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Background

This vignette goes through the steps involved in refactoring the
function `forward` to increase performance.

# Setup

We'll use the following notation

+ `T` = total number of time points ($T=1$ in this case)

+ `K` = number of states (2 in this case)

+ `D` = dimension of observations ($D>1$ is multivariate) ($D=1$ in this case)

+ `N` = number of participants ($N=69$ in this case)

## Test Data

The test data is included in this package, it's called `test_data`.
It's a list with the following members:

+ `A` = $T \times K \times K$ array with transition probabilities,
  from row to column.  In practice it ends up being a $K \times K$
  matrix.

+ `B` = $T \times D \times K$ matrix with elements $ab_{ij} = P(y_i|s_j)$.

+ `init` = $N \times K$ matrix with initial probabilities.

## Profiling the original code

```{r,echo=F,warning=F,message=F}
library(tidyverse)
library(hmmfb)
library(profvis)
library(microbenchmark)
```

We'll start by using `profvis` to find hotspots in the original code.

```{r,eval=F}
data("test_data")
A <- test_data$A
B <- test_data$B
ntimes <- test_data$ntimes
init <- test_data$init
mprof <- profvis(forward_old(init,A,B,ntimes=ntimes,return.all = T))
mprof
```

```{r,echo=F}
data("profvis_dat")
profvis_dat
```

The profiling revealed that the hotspot is on line `106`.How much of
an improvement do we get if we replace that line with a call to a pure
C++ function? I'll use the package `microbenchmark` to time the
results (each function will be called 10 times). Results are displayed
as relative runtime compared to the best median time.

```{r,echo=F}
data("first_bench")
first_bench
```

Looks like the swapping out that one function has given about a *7x*
speedup. Pretty good for writing 5 lines of code!

## Faster still

What if we write the whole function in C++? 

```{r,echo=F}
data("second_bench")

second_bench
```

The pure C++ function is a whopping *115* times faster than the
original function, and about 15x faster than the already optimized
code. Pretty good!
